---
title: "bif_rda_analysis"
author: "ruben"
date: "09/06/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Librairies

```{r,message=F,warning=F}

library(dplyr)
library(readr)
library(pheatmap)
library(data.table)
library(tidyr)
library(tibble)
library("factoextra")
library("FactoMineR")
library(RColorBrewer)
library(tidyverse)
library(broom)


source(file = "functions.R")

```

## import data

```{r,message=F,warning=F}

load("bif_mags_dist.rda")

# name : bif_mags_dist

colnames(bif_mags_dist) <- gsub("\\.fa", "", colnames(bif_mags_dist))
  
rownames(bif_mags_dist) <- gsub("\\.fa", "", rownames(bif_mags_dist))

load("curated_v3_otu_tax.rda")

mags_tax_gene_id_eggnog <- readr::read_csv2("data-raw/bif_mags/mags_tax_gene_id_eggnog.csv") %>% 
   select(-1) %>%
  ## filter the species of interest
  filter(grepl("adolescentis|animalis| longum|bifidum|dentium|pseudocatenulatum|catenulatum|animalis", assigned_species)) %>%
  ### eggnog is linked to several annotations KEGG, CAZy, so we link only 1 annotations for 1 eggnog
  group_by(eggNOG_OGs) %>%
  mutate(KEGG_ko = get_mode(KEGG_ko)) %>%
  mutate(CAZy = get_mode(CAZy)) %>%
  mutate(EC = get_mode(EC)) %>%
  ungroup() %>%
  filter(completeness > 80)



mags_tax_file = readr::read_tsv("data-raw/bif_mags/Bifidobacterium_annotation/annotations_url_metadata_opendata.tsv") %>%
  filter(assigned_genus=="Bifidobacterium") %>%
  filter(completeness > 80) %>%
  filter(grepl("adolescentis|animalis| longum|bifidum|dentium|pseudocatenulatum|catenulatum|animalis", assigned_species)) %>%
  select(genome_name,study,sample_name,assigned_species,completeness, average_distance)

sample_data <- sampleMetadata

## if one ppl has several samples, keep the sample with the highest nb of reads
sample_unique <- unique(setDT(sample_data)[order(subject_id, -number_reads)], by = "subject_id")

sample <- sample_unique %>%
   ## transform disease and westernized to good format
   transform_disease(.) %>%
   transform_westernized(.) %>%
   select(sample_id, body_site, disease, age_category, gender, country, westernized, BMI, antibiotics_current_use)

```

## prepare annotation table

```{r,message=F,warning=F}

bif_mags_dist_annotations <- mags_tax_file %>%
  select(genome_name, sample_name, assigned_species) %>%
  merge(., sample %>% select(sample_id, disease, age_category, westernized) , by.x = "sample_name", by.y = "sample_id") %>%
  distinct()

```

## select only species of interest + completeness > 80%

```{r,message=F,warning=F}

unique_mags <- unique(bif_mags_dist_annotations$genome_name)

bif_mags_dist_dt <- bif_mags_dist %>%
  select(matches(unique(bif_mags_dist_annotations$genome_name))) %>%
  filter(rownames(.) %in% unique(bif_mags_dist_annotations$genome_name)) %>%
  select(rownames(.))

##reorder columns as rownames
bif_mags_dist_dt <- bif_mags_dist_dt[,rownames(bif_mags_dist_dt)]


```


## filter the data to check the high distances genomes, delete the high distance genome

```{r,message=F,warning=F, fig.height=7, fig.width = 8}

bif_mags_dist_dt_filtered <- bif_mags_dist_dt %>%
  rownames_to_column("mags_a") %>%
  gather("mags_b","distance", -1) %>%
  filter(distance > 0.8)

bif_mags_dist_dt_filtered

mags_to_delete <- union(bif_mags_dist_dt_filtered$mags_a,bif_mags_dist_dt_filtered$mags_b)



```


## plot pheatmap

```{r,message=F,warning=F, fig.height=7, fig.width = 8}


bif_mags_ft <- bif_mags_dist_dt %>%
  select(-mags_to_delete ) %>%
  filter(!row.names(.) %in% mags_to_delete)

pheatmap(bif_mags_ft,
         show_colnames = FALSE, show_rownames = FALSE,
         annotation_row = bif_mags_dist_annotations %>%
           #select(-sample_name) %>% column_to_rownames("genome_name"),
           select(assigned_species, genome_name) %>% column_to_rownames("genome_name"),
         annotation_col = bif_mags_dist_annotations %>%
           #select(-sample_name) %>% column_to_rownames("genome_name"))
           select(assigned_species, genome_name) %>% column_to_rownames("genome_name"))
```


## plot pheatmap per species

```{r,message=F,warning=F, fig.height=7, fig.width = 8}


species <- c("Bifidobacterium adolescentis","Bifidobacterium animalis","Bifidobacterium bifidum","Bifidobacterium catenulatum","Bifidobacterium longum","Bifidobacterium pseudocatenulatum")


for (one_species in species){
  
  ## which mags are assigned to one_species ?
  
  mags_assigned_to_one_species <- bif_mags_dist_annotations %>%
    filter(assigned_species == one_species) %>%
    .$genome_name
  
  
  bif_mags_one_species <- bif_mags_dist_dt %>%
    select(mags_assigned_to_one_species) %>%
    filter(row.names(.) %in% mags_assigned_to_one_species)
  
  pheatmap(bif_mags_one_species,
           main = paste0(one_species),
         show_colnames = FALSE, show_rownames = FALSE,
         annotation_row = bif_mags_dist_annotations %>%
           select(-sample_name, -assigned_species, -westernized) %>% column_to_rownames("genome_name"),
           #select(assigned_species, genome_name) %>% column_to_rownames("genome_name"),
         annotation_col = bif_mags_dist_annotations %>%
           select(-sample_name, -assigned_species, -westernized) %>% column_to_rownames("genome_name"))
           #select(assigned_species, genome_name) %>% column_to_rownames("genome_name"))
  
}


```

## plot pheatmap per species

```{r eval=FALSE, fig.height=7, fig.width=, include=FALSE}


mags_tax_not_filtered <- readr::read_tsv("data-raw/bif_mags/Bifidobacterium_annotation/annotations_url_metadata_opendata.tsv") %>%
  #ungroup() %>%
  group_by(sample_name) %>%
  count(assigned_species)

max(mags_tax_not_filtered$n)

```


## prepare bifidum mash heatmap

```{r,message=F,warning=F, fig.height=7, fig.width = 8}

one_species <- "Bifidobacterium bifidum"

## are the clusters observed in bifidum associated with the phage signal observed differentially in
## healthy/unhealthy bifidotypes ?

## load bifidotypes file

bifido <- fread("enterotype_DMM") %>%
   select(-V1) %>%
   mutate(grp_healthy = case_when(
    bifidotype %in% c(1,2,6) ~ "healthy",
    TRUE ~ "unhealthy"
  ))

## are the clusters observed in bifidum associated with the phage signal differentially present in
## healthy/unhealthy bifidotypes ?
bif_mags_dist_annotations <- mags_tax_file %>%
  select(genome_name, sample_name, assigned_species) %>%
  merge(., sample %>% select(sample_id, disease, age_category, westernized) , by.x = "sample_name", by.y = "sample_id") %>%
  mutate(disease = case_when(disease %in% c("Control") ~ "healthy",TRUE ~ "unhealthy")) %>%
  merge(bifido %>% select(-bifidotype), by.x = "sample_name", by.y = "sample_id", all.x = TRUE) %>%
  distinct()
  
  

mags_assigned_to_one_species <- bif_mags_dist_annotations %>%
  filter(assigned_species == one_species) %>%
  .$genome_name
  
  
bif_mags_one_species <- bif_mags_dist_dt %>%
  select(mags_assigned_to_one_species) %>%
  filter(row.names(.) %in% mags_assigned_to_one_species)

```

## bifidum mash heatmap

```{r,message=F,warning=F, fig.height=7, fig.width = 8}


test <- bif_mags_dist_annotations %>%
  select(-sample_name, -assigned_species, -westernized, -disease, -age_category) %>%
  column_to_rownames("genome_name")

pheatmap(bif_mags_one_species,
         clustering_method = "ward.D",
         main = paste0(one_species),
         show_colnames = FALSE, show_rownames = FALSE,
         annotation_row = bif_mags_dist_annotations %>%
           select(-sample_name, -assigned_species, -westernized, -disease, -age_category) %>% column_to_rownames("genome_name"),
           #select(grp_healthy, genome_name) %>% column_to_rownames("genome_name"),
           #select(assigned_species, genome_name) %>% column_to_rownames("genome_name"),
         annotation_col = bif_mags_dist_annotations %>%
           #select(grp_healthy, genome_name)) %>% column_to_rownames("genome_name"))
           select(-sample_name, -assigned_species, -westernized, -disease, -age_category) %>% column_to_rownames("genome_name"))
           #select(assigned_species, genome_name) %>% column_to_rownames("genome_name"))

## eggnogs with p.value < 1%, which means they are differentially present/absent in healthy/unhealthy
## subjects (bifidotypes) in bifidum mags

```


## bifidum mash heatmap

```{r,message=F,warning=F, fig.height=7, fig.width = 8}

## eggnogs with p.value < 1%, which means they are differentially present/absent in healthy/unhealthy
## subjects (bifidotypes) in bifidum mags

eggnogs_related_to_phages_w <- c("COG0582@1|root,COG0582@2|Bacteria,2HZMA@201174|Actinobacteria,4D0X7@85004|Bifidobacteriales",
"COG0582@1|root,COG0732@1|root,COG0582@2|Bacteria,COG0732@2|Bacteria,2HUXN@201174|Actinobacteria,4CZ21@85004|Bifidobacteriales",
"28IBR@1|root,2Z8E5@2|Bacteria,2IF1B@201174|Actinobacteria,4D2Y2@85004|Bifidobacteriales",
"COG4926@1|root,COG4926@2|Bacteria,2GKQD@201174|Actinobacteria,4CZ3Q@85004|Bifidobacteriales",
"COG3941@1|root,COG5412@1|root,COG3941@2|Bacteria,COG5412@2|Bacteria,2H75F@201174|Actinobacteria,4CZ35@85004|Bifidobacteriales")


eggnogs_related_to_phages<- c("28IBR@1|root,2Z8E5@2|Bacteria,2IF1B@201174|Actinobacteria,4D2Y2@85004|Bifidobacteriales",
"2B2EW@1|root,31UZI@2|Bacteria,2GWDP@201174|Actinobacteria,4D1SW@85004|Bifidobacteriales",
"2B5PP@1|root,31YJ5@2|Bacteria,2I00C@201174|Actinobacteria,4D2GW@85004|Bifidobacteriales",
"2B5R3@1|root,31YKM@2|Bacteria,2H0XN@201174|Actinobacteria,4D2J8@85004|Bifidobacteriales",
"2EINA@1|root,30Z4H@2|Bacteria,2I8AV@201174|Actinobacteria,4CYPM@84998|Coriobacteriia",
"arCOG12631@1|root,31Y2B@2|Bacteria,2HZMF@201174|Actinobacteria,4D0XK@85004|Bifidobacteriales",
"COG0616@1|root,COG0616@2|Bacteria,1MY1I@1224|Proteobacteria,42PFB@68525|delta/epsilon subdivisions,2WSJ6@28221|Deltaproteobacteria",
"COG1073@1|root,COG1073@2|Bacteria,2HZEB@201174|Actinobacteria,4D02D@85004|Bifidobacteriales",
"COG1086@1|root,COG2148@1|root,COG1086@2|Bacteria,COG2148@2|Bacteria,2GK0M@201174|Actinobacteria,4CZ4J@85004|Bifidobacteriales",
"COG1670@1|root,COG1670@2|Bacteria,2HZK6@201174|Actinobacteria,4D0RS@85004|Bifidobacteriales",
"COG2023@1|root,COG2023@2|Bacteria,2GVUQ@201174|Actinobacteria,4D1JT@85004|Bifidobacteriales",
"COG2801@1|root,COG2801@2|Bacteria,2GKDY@201174|Actinobacteria,4CW9S@84998|Coriobacteriia",
"COG2963@1|root,COG2963@2|Bacteria,2HV7B@201174|Actinobacteria,4CWFM@84998|Coriobacteriia",
"COG3378@1|root,COG3378@2|Bacteria,2IC10@201174|Actinobacteria,4D0F0@85004|Bifidobacteriales",
"COG3677@1|root,COG3677@2|Bacteria,2IEJD@201174|Actinobacteria,4CZTH@85004|Bifidobacteriales"
)




## calculate the prevalence of these eggnogs in mags

test <- mags_tax_gene_id_eggnog %>%
  select(genome_name, eggNOG_OGs) %>%
  filter(eggNOG_OGs %in% eggnogs_related_to_phages) %>%
  group_by(genome_name, eggNOG_OGs) %>%
  distinct() %>%
  summarise(n=1) %>%
  reshape2::dcast(eggNOG_OGs~genome_name, value.var = "n", fill=0) %>%
  column_to_rownames("eggNOG_OGs") %>%
  gather("genome_name", "presence") %>%
  ungroup() %>%
  group_by(genome_name) %>%
  summarise(prevalence_phage = (sum(presence)/n())) %>%
  ## add bifidotypes
  merge(bif_mags_dist_annotations, by="genome_name", all.x = TRUE) %>%
  select(genome_name, grp_healthy, prevalence_phage) %>%
  column_to_rownames("genome_name")


```

## bifidum mash heatmap

```{r,message=F,warning=F, fig.height=5, fig.width = 8}

test %>%
  dplyr::rename(`community types` = grp_healthy, `differential OGs prevalence` = prevalence_phage) %>%


pheatmap(bif_mags_one_species,
         clustering_method = "ward.D",
         #main = paste0(one_species),
         show_colnames = FALSE, show_rownames = FALSE, annotation_names_row = FALSE,
         annotation_row = .) 

ggsave("bif_bifidum_mags.pdf")

```


```{r,message=F,warning=F, fig.height=5, fig.width = 8}




one_species <- "Bifidobacterium bifidum"

## are the clusters observed in bifidum associated with the phage signal observed differentially in
## healthy/unhealthy bifidotypes ?

## load bifidotypes file

bifido <- fread("enterotype_DMM") %>%
   select(-V1) %>%
   mutate(grp_healthy = case_when(
    bifidotype %in% c(1,2,6) ~ "healthy",
    TRUE ~ "unhealthy"
  ))

## are the clusters observed in bifidum associated with the phage signal differentially present in
## healthy/unhealthy bifidotypes ?
bif_mags_dist_annotations <- mags_tax_file %>%
  select(genome_name, sample_name, assigned_species) %>%
  merge(., sample %>% select(sample_id, disease, age_category, westernized) , by.x = "sample_name", by.y = "sample_id") %>%
  mutate(disease = case_when(disease %in% c("Control") ~ "healthy",TRUE ~ "unhealthy")) %>%
  merge(bifido %>% select(-bifidotype), by.x = "sample_name", by.y = "sample_id", all = FALSE) %>%
  distinct()
  
  

mags_assigned_to_one_species <- bif_mags_dist_annotations %>%
  filter(assigned_species == one_species) %>%
  .$genome_name
  
  
bif_mags_one_species <- bif_mags_dist_dt %>%
  select(mags_assigned_to_one_species) %>%
  filter(row.names(.) %in% mags_assigned_to_one_species)


test <- mags_tax_gene_id_eggnog %>%
  select(genome_name, eggNOG_OGs) %>%
  filter(eggNOG_OGs %in% eggnogs_related_to_phages) %>%
  group_by(genome_name, eggNOG_OGs) %>%
  distinct() %>%
  summarise(n=1) %>%
  reshape2::dcast(eggNOG_OGs~genome_name, value.var = "n", fill=0) %>%
  column_to_rownames("eggNOG_OGs") %>%
  gather("genome_name", "presence") %>%
  ungroup() %>%
  group_by(genome_name) %>%
  summarise(prevalence_phage = (sum(presence)/n())) %>%
  ## add bifidotypes
  merge(bif_mags_dist_annotations, by="genome_name", all.x = TRUE) %>%
  select(genome_name, grp_healthy, prevalence_phage) %>%
  column_to_rownames("genome_name")

test %>%
  dplyr::rename(`community types` = grp_healthy, `differential OGs prevalence` = prevalence_phage) %>%
  filter(!is.na(`community types`)) %>%


pheatmap(bif_mags_one_species,
         clustering_method = "ward.D",
         #main = paste0(one_species),
         show_colnames = FALSE, show_rownames = FALSE, annotation_names_row = FALSE,
         annotation_row = .) 

ggsave("bif_bifidum_mags_2.pdf")

```


## predict the bifidotypes of ppl which don't have bifidotypes in metadata

```{r,message=F,warning=F, fig.height=7, fig.width = 8}

load("curated_v3_otu_tax.rda")

otu_full <- OTU %>%
  data.frame(.)

tax_full <- TAX %>%
  data.frame(.)

sample <- sampleMetadata %>%
  select(sample_id, subject_id, antibiotics_current_use, disease, body_site, age_category, non_westernized)

### import DMM model based on 6 clusters

```

## predict the bifidotypes of ppl which don't have bifidotypes in metadata

```{r,message=F,warning=F, fig.height=7, fig.width = 8}

## load best_fit from dmm and bifidotypes

load("DMM_files/best_fit_DMM_6clusters")

bifido <- fread("enterotype_DMM") %>%
   select(-V1)

## filter the count table

count <- otu_full %>%
  ## filter bifid species only
  filter(row.names(.) %in% (tax_full %>% filter(Genus == "Bifidobacterium") %>% rownames(.))) %>%
  ## select ppl who are not used in the learning set
  select(-bifido$sample_id) %>%
  t() %>%
  as.matrix()

# Pick the OTU count matrix
# and convert it into samples x taxa format
dat <- count
count_raw <- as.matrix(t(dat))

# rowsums equal zero (sample with 0 count everywhere) need to be deleted

sample_low_reads <- names(which(rowSums(count_raw) <= 500))


## select ppl with 0 bifid, they are selected as a no bifid cluster

no_bifid <- count_raw %>%
  colSums(.) %>%
  data.frame(sum = .) %>%
  filter(sum==0) %>%
  rownames(.)

# filter the count table

count <- count_raw %>%
  as.data.frame(.) %>%
  select(-no_bifid)

# colsums equal zero (bacteria not present in any sample) need to be deleted

otu_not_present <- names(which(rowSums(count) == 0))

count <- count %>%
  filter(!rownames(.) %in% otu_not_present) %>%
  ## filtering ppl with nb reads < threshold
  filter(!row.names(.) %in% sample_low_reads) %>%
  as.matrix(.)

# transform colnames to species name

rownames(count) <- get_species(as.data.frame(tax_full),rownames(count))

load("./DMM_files/best_fit_DMM_6clusters")

nrow(count)


tax_full %>%
  filter(Genus == "Bifidobacterium") %>%
  rownames(.)

```

## process for predicting bifidotypes

```{r,message=F,warning=F, fig.height=7, fig.width = 8}
## we need to select the common species in fit$estimate


common_species <- intersect(rownames(best@fit$Estimate), rownames(count))

## filter according to common species
count <- count %>%
  as.data.frame %>%
  filter(row.names(.) %in% common_species)



best_fit <- best


bifidotypes_prediction = predict(best, t(count) , assign = TRUE)



####

best_fit <- best

count_learning_set <- count %>%
  t %>%
  as.data.frame

## we need to select the common species between count and best fit object (estimate)
best_fit@fit$Estimate <- best_fit@fit$Estimate %>% as.data.frame %>% filter(row.names(.) %in% common_species) %>% as.matrix


predict_df <- predict(best_fit, count_learning_set, assign=TRUE) %>%
  as.data.frame


colnames(predict_df) <- 1:6

## create a column which will check which bifidotype corresponds to each subject (maximum of fit score per row)

predict_df$assigned_bifidotype = colnames(predict_df)[apply(predict_df, 1, which.max)]


bifido_other_sample <- predict_df %>%
  select(assigned_bifidotype) %>%
  dplyr::rename(bifidotype = assigned_bifidotype) %>%
  rownames_to_column("sample_id")


## we will add this to the bifidotypes file and create a full bifidotype on "all" samples (samples which are unique + most reads)

bifido_whole <- bifido_other_sample %>%
  rbind(bifido)


### we need to add the ppl who are part of the bifidotypes 7 (no bifidobacterium reads)

cluster_7 <- colnames(otu_full) %>%
  data.frame(sample_id = .) %>%
  filter(!sample_id %in% bifido_whole$sample_id) %>%
  mutate(bifidotype = 7)

bifido_whole <- bifido_whole %>%
  rbind(cluster_7)


```

## process for predicting bifidotypes

```{r,message=F,warning=F, fig.height=7, fig.width = 8}

### we will check the number of each bifidotype assigned

predict_df %>%
  select(assigned_bifidotype) %>%
  group_by(assigned_bifidotype) %>%
  summarise(no_rows = length(assigned_bifidotype))


```


## we will add the predicted values so that the heatmap is completed !

```{r,message=F,warning=F, fig.height=7, fig.width = 8}

## classify as healthy/unhealthy bifidotypes

bifido_whole <- bifido_whole %>%
   mutate(grp_healthy = case_when(
    bifidotype %in% c(1,2,6) ~ "healthy",
    TRUE ~ "unhealthy"
  ))

### we will complete the bif_mags_dist_annotations file with new bifidotypes

bif_mags_dist_annotations_completed <- bif_mags_dist_annotations %>%
  select(-grp_healthy) %>%
  merge(bifido_whole %>% dplyr::rename(sample_name = sample_id), by = "sample_name")


### prepare heatmap annotation


test <- mags_tax_gene_id_eggnog %>%
  select(genome_name, eggNOG_OGs) %>%
  filter(eggNOG_OGs %in% eggnogs_related_to_phages) %>%
  group_by(genome_name, eggNOG_OGs) %>%
  distinct() %>%
  summarise(n=1) %>%
  reshape2::dcast(eggNOG_OGs~genome_name, value.var = "n", fill=0) %>%
  column_to_rownames("eggNOG_OGs") %>%
  gather("genome_name", "presence") %>%
  ungroup() %>%
  group_by(genome_name) %>%
  summarise(prevalence_phage = (sum(presence))) %>%
  ## add bifidotypes
  merge(bif_mags_dist_annotations_completed, by="genome_name", all.x = TRUE) %>%
  select(genome_name, grp_healthy, prevalence_phage) %>%
  column_to_rownames("genome_name")

test$prevalence_phage <- as.numeric(test$prevalence_phage)


```

## we will add the predicted values so that the heatmap is completed !

```{r,message=F,warning=F, fig.height=4, fig.width = 7}

colnames(test) <- c("bifidotype", "selected EggnOGs presence")

pheatmap(bif_mags_one_species %>%
           filter(row.names(.) %in% rownames(test)) %>%
           select(matches(rownames(test))),
         clustering_method = "ward.D2",
         main = paste0(one_species),
         show_colnames = FALSE, show_rownames = FALSE, annotation_names_row = FALSE,
         annotation_row = test)
         #annotation_col = test)

```
