---
title: "MAGS_analysis"
author: "ruben"
date: "22/04/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Librairies

```{r,message=F,warning=F}

library(dplyr)
library(readr)
library(pheatmap)
library(data.table)
library(tidyr)


source(file = "functions.R")

```

## Données

```{r,message=F,warning=F}

load("curated_v3_otu_tax.rda")

## MAGS annotations data
mags_tax_file = readr::read_tsv("MAGS_files/annotations_url_metadata_opendata.tsv") %>%
  filter(assigned_genus=="Bifidobacterium") %>%
  select(genome_name,study,sample_name,assigned_species,completeness, average_distance)

# MAGS linked to other ID's
mags_tax_gene_id_eggnog <- readr::read_csv2("MAGS_files/mags_tax_gene_id_eggnog.csv") %>%
   select(-X1)

## number of mags per species

nb_mags_species <- mags_tax_file %>%
  data.frame(.) %>%
  group_by(assigned_species) %>%
  count() %>%
  filter(!assigned_species %in% c("Bifidobacterium sp", "N/A"))
nb_mags_species$assigned_species <- gsub("Bifidobacterium","B.",nb_mags_species$assigned_species)

```

## Plot the number of mags per species of bifidobacterium

```{r,message=F,warning=F, fig.height=7, fig.width = 12}

ggplot(nb_mags_species, aes(x = assigned_species, y = n, fill = assigned_species)) +  # Plot with values on top
  geom_bar(stat = "identity") +
  #geom_bar(width = 1, stat = "identity", color = "white") +
  #coord_polar("y", start = 0)+
  geom_text(aes(label = n), vjust = 0) +
  scale_x_discrete(guide = guide_axis(n.dodge=2)) +
  labs( y = "number of MAGS", x = "", fill = "Bifidobacterium assigned species")

```

## Choose the functionnality (Eggnog, Cazy, etc)

```{r,message=F,warning=F}

which_function <- readline(prompt=paste("Which functions you want to plot ? : ", "Kegg-KO - ko", "Cazy - ca","EC - ec","Eggnog - eg","", sep="\n"))

while(!which_function %in% c("ko", "ca", "ec", "eg")){
  which_function <- readline(prompt=paste("Error, please redo : ", "Kegg-KO - ko", "Cazy - ca","", sep="\n"))
}

if(which_function == "ko"){
  ## we exclude predict data
  
  functionality <- "KEGG_ko"
  
} else if(which_function == "ca"){
  ## we keep only predict data
  
  functionality <- "CAZy"
  
}else if(which_function == "ec"){
  ## we keep only predict data
  
  functionality <- "EC"
  
}else if(which_function == "eg"){
  ## we keep only predict data
  
  functionality <- "seed_ortholog"
  
}


```

## Prepare plot

```{r,message=F,warning=F}


mags_ko_table <- mags_tax_gene_id_eggnog %>%
  filter(completeness >= 80) %>%
  filter((!!sym(functionality)) != "-") %>%
  group_by_at(functionality) %>%
  mutate(ko_n=n()) %>%
   filter(ko_n>10) %>%
   ungroup() %>%
   group_by_at(vars('genome_name',functionality)) %>%
   summarise(n=n()) %>%
   mutate(n=ifelse(n>1,2,n)) %>%
   reshape2::dcast(paste0('genome_name', '~', functionality), value.var = "n", fill=0)



heat <- mags_ko_table %>%
  tibble::column_to_rownames("genome_name") %>%
  as.matrix()

```

## filter heatmap/add metadata

```{r,message=F,warning=F}

sample <- sampleMetadata

## if one ppl has several samples, keep the sample with the highest nb of reads
sample <- unique(setDT(sampleMetadata)[order(subject_id, -number_reads)], by = "subject_id")

metadata <- sample %>%
   ## transform disease and westernized to good format
   transform_disease(.) %>%
   transform_westernized(.) %>%
   select(sample_id, body_site, disease, age_category, gender, country, westernized, BMI, antibiotics_current_use)


## merge with mags_tax

mags_tax <- mags_tax_file %>%
   merge(metadata, by.x = "sample_name", by.y = "sample_id") %>%
   tibble::column_to_rownames("genome_name") %>%
   filter(assigned_species != "N/A") %>%
   filter(grepl("adolescentis|animalis|longum|bifidum|dentium|pseudocatenulatum|catenulatum|animalis", assigned_species))

```

## Filter heat

```{r,message=F,warning=F}

## we filter heat lines which are in mags_tax rownames

heat <- heat %>%
   data.frame(.) %>%
   filter(row.names(.) %in% rownames(mags_tax))


# the species
species <- levels(factor(mags_tax$assigned_species))

```

## Delete the species not assigned to MAGS

```{r,message=F,warning=F, fig.height=9, fig.width = 9}

for (one_species in species){
   
   mags_tax_filtered <- mags_tax %>% filter(assigned_species == one_species)
   
   heat_filtered <- heat %>% filter(row.names(.) %in% rownames(mags_tax_filtered))
   
   
   if(nrow(heat_filtered) <10){
      
      species <- species[!species %in% one_species]
      
      }
      
   }

mags_tax <- mags_tax %>% filter(assigned_species %in% species)

```

## Plotting cazy enzymes

```{r,message=F,warning=F, fig.height=9, fig.width = 9}

library("remotes")
library("RColorBrewer")

palette <- brewer.pal(n = 12, name = 'Paired')


## number of species to respresent
nb_species <- length(species)

annotation_colors = list(
  assigned_species = palette[1:nb_species])

## assign names to the list of colors

names(annotation_colors$assigned_species) <- levels(factor(mags_tax$assigned_species))

```


## If functions are eggnog, load the cluster file

```{r,message=F,warning=F, fig.height=9, fig.width = 9}


rep <- "./MAGS_files/"
file_name <- "clusterboot"


if(!file_name %in% list.files(rep)){
  
  print("please run MAGS_clustering_analysis.Rmd to print the eggnog clusters")
  
}else{
  load(paste0(rep, file_name))
}

```

## Plotting one MAGS per species

```{r,message=F,warning=F, fig.height=9, fig.width = 9}

## if the annotation is based on seed_ortholog, just print Eggnog in the plot

if (which_function == 'eg'){
  functionality <- "EGGnog"
}

## plot a mags per species

if(which_function != "eg"){
  
  
  for (one_species in species & which_function != "eg"){
   
    mags_tax_filtered <- mags_tax %>% filter(assigned_species == one_species)
   
    pheat <- pheatmap(heat %>% filter(row.names(.) %in% rownames(mags_tax_filtered)),
                      #cutree_rows = 9,
                      show_rownames = FALSE,
                      show_colnames = FALSE,
                      annotation_row = mags_tax_filtered %>%
                        select(disease, westernized, age_category),
                      annotation_colors = annotation_colors,
                      main= paste0(one_species, " MAGS | ", functionality, " annotations")
                      )

    pheat
    }

  }

```

## Plot eggnogs only

```{r,message=F,warning=F, fig.height=9, fig.width = 9}

## define threshold to filter eggnog based on their prevalence

threshold = 0.10

## if clustering has been done, add a column panel showing clusters

if("cluster_results" %in% list.files(rep) & which_function == "eg"){
  
  for (one_species in species){
    
    # for each species, select mags
    mags_tax_filtered <- mags_tax %>% filter(assigned_species == one_species)


    heat_filtered <-  heat %>%
      filter(row.names(.) %in% rownames(mags_tax_filtered)) %>%
      select(rownames(cluster_results))
    
    annot_col <- heat_filtered
    rownames(annot_col) <- NULL
    
    annot_col <- annot_col %>%
      summarise_all(., funs(mean(.>0))) %>%
      t(.) %>%
      data.frame(prevalence = .) %>%
      rownames_to_column("eggnog")
      
    
    ### which eggnog we delete ? based on a theshold of prevalence (% ?)
    
    eggnog_filtered <- MAGS_eggnog_prevalence %>%
      filter(prevalence > threshold & prevalence < 1 - threshold ) %>%
      .$eggnog %>% as.character(.)
    
    annot_filtered <- heat_filtered %>%
      select(eggnog_filtered) %>%
      t %>%
      data.frame(.)
    
    
      clusters_msp <- annot_filtered %>%
        ## need to select the distinct row to calculate the dissimilarity date
        distinct(.) %>%
    ade4::dist.binary(.,method=1) %>%
    fpc::clusterboot(.,B=100,
                     bootmethod="subset",
                     clustermethod=fpc::pamkCBI,
                     count=FALSE,
                     k=1:6, showplot=FALSE,
                     seed = seed)
      
      
      annot_final <- clusters_msp$result$partition %>%
        data.frame(cluster = .)
      annot_final$cluster <- as.character(annot_final$cluster)
      
                        
    ### plot the phetmap
    
    
   
   mags_tax_filtered <- mags_tax %>% filter(assigned_species == one_species)
   
   pheat <- pheatmap(heat_filtered,
                     #cutree_rows = 9,
                     show_rownames = FALSE,
                     show_colnames = FALSE,
                     annotation_row = mags_tax_filtered %>%
                        select(disease, westernized, age_category),
                     annotation_colors = annotation_colors,
                     main= paste0(one_species, " MAGS | ", functionality, " annotations"),
                     annotation_col = annot_final
                     )

   pheat
  }
  
}

```


## Prepare heatmap of functions prevalence per species

```{r,message=F,warning=F, fig.height=9, fig.width = 9}


  
mags_tax_filtered <- mags_tax %>%
  rownames_to_column("genome_name") %>%
  select(genome_name, assigned_species)
  
mags_function_table <- mags_ko_table %>%
  filter(genome_name %in% mags_tax_filtered$genome_name) %>%
  ## merge with mags_tax_filtered to have the completeness critera and assigned species
  merge(mags_tax_filtered, by="genome_name") %>%
  column_to_rownames("genome_name") %>%
  ### al positive value become 1 for prevalennce count
  mutate_at(vars(-("assigned_species")),funs(ifelse(.>0,1,.))) %>%
  ## group_by species
  group_by(assigned_species) %>%
  summarise_all(., funs(sum(.)/n()))
  
  
  

```


## Plot heatmap

```{r,message=F,warning=F, fig.height=9, fig.width = 9}

mags_heatmap <- mags_function_table %>%
  column_to_rownames("assigned_species")

pheatmap(mags_heatmap,
                     #show_rownames = FALSE,
                     show_colnames = FALSE,
                     )

```